---
title: "My Project"
subtitle: "Small Business Road to Success"
author: "LaVelle White"
institute: "Chaminade University"
date: "04/14/25"
format:
  revealjs:
    theme: sky
    transition: slide
    slide-number: true
editor: visual
---

# Introduction

This project explores how different revenue levels predict key indicators of small business success using survey data from the Small Business Credit Survey.

------------------------------------------------------------------------

# Packages Used

```{r packages, echo=TRUE}
library(tidyverse)     # Core data wrangling & plotting
library(janitor)       # Clean column names
library(readxl)        # Read Excel files
library(ggplot2)       # Data visualization
```

------------------------------------------------------------------------

# Data Source

-   **Dataset**: 2024 Small Business Credit Survey (Employer Firms)\
-   **File Name**: `sbcs-employer-firms-appendix-2024.xlsx`\
-   **Sheet Used**: `Revenue`\
-   **Provider**: [Federal Reserve Banks](https://www.fedsmallbusiness.org)

------------------------------------------------------------------------

# Objective

To investigate how revenue levels predict various indicators of small business success using data from the *2024 Small Business Credit Survey*.

::: small
I used logistic regression models to assess relationships between revenue and the following outcomes:

1.  Receiving loans or grants\
2.  Applying for financing\
3.  Receiving most/all requested funding\
4.  Operating at a profit\
5.  Reporting strong financial condition\
6.  Expecting employment growth\
7.  A combined success model
:::

```{r read-data, echo=TRUE}
revenue_data <- read_excel("C:/Users/12103/OneDrive - Chaminade University of Honolulu/Documents/DS303_SP25/sbcs-employer-firms-appendix-2024.xlsx", sheet = "Revenue")
```

------------------------------------------------------------------------

# Approach

-   **Source**: 2024 Fed Small Business Credit Survey (Revenue Sheet)
-   **Method**: Logistic regression (binomial family)
-   **Metrics**: AIC scores and ANOVA p-values to evaluate model strength

------------------------------------------------------------------------

# Data Preparation

```{r data-cleaning, include=TRUE}
clean_names <- revenue_data %>% clean_names()
clean_names <- clean_names %>%
  mutate(percent = parse_number(as.character(percent)) / 100)
```

------------------------------------------------------------------------

# Model 1: Loan or Grant Received (Setup)

```{r model1, include=TRUE}
actions_success <- clean_names %>%
  filter(str_detect(survey_question, "(?i)actions taken in response to financial challenges")) %>%
  mutate(success = case_when(
    str_detect(response_option, "must be repaid") ~ 1,
    str_detect(response_option, "do not have to be repaid") ~ 1,
    TRUE ~ 0
  ),
  yes = round(percent * question_sample_size),
  no = question_sample_size - yes,
  revenue = factor(revenue)) %>%
  group_by(revenue, success) %>%
  summarise(yes = sum(yes), no = sum(no), .groups = "drop")

model_actions_success <- glm(cbind(yes, no) ~ revenue, data = actions_success, family = binomial())
actions_success$predicted <- predict(model_actions_success, type = "response")
```

------------------------------------------------------------------------

# Model 1: Loan or Grant Received (Plot)

```{r plot1, echo=FALSE}
ggplot(actions_success, aes(x = revenue, y = predicted)) +
  geom_col(fill = "#377eb8") +
  coord_flip() +
  labs(
    title = "Predicted Probability of Receiving Loan or Grant",
    x = "Revenue Group",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

# Actions Taken During Financial Challenges

-   AIC: 697.69\
-   p \< 2.2e-16\
-   **Insight**: High revenue businesses (especially \$5M+) were less likely to rely on external funds. Smaller firms depended more on loans/grants during hardship.

------------------------------------------------------------------------

# Model 2: Applied for Financing (Setup)

```{r model2, include=TRUE}
application_success <- clean_names %>%
  filter(str_detect(survey_question, "(?i)applications for financing")) %>%
  mutate(success = if_else(response_option == "Applied for a loan, line of credit, or merchant cash advance", 1, 0),
         yes = round(percent * question_sample_size),
         no = question_sample_size - yes,
         revenue = factor(revenue)) %>%
  group_by(revenue, success) %>%
  summarise(yes = sum(yes), no = sum(no), .groups = "drop")

model_applications <- glm(cbind(yes, no) ~ revenue, data = application_success, family = binomial())
application_success$predicted <- predict(model_applications, type = "response")
```

------------------------------------------------------------------------

# Model 2: Applied for Financing (Plot)

```{r plot2, echo=FALSE}
ggplot(application_success, aes(x = revenue, y = predicted)) +
  geom_col(fill = "#4daf4a") +
  coord_flip() +
  labs(
    title = "Predicted Probability of Applying for Financing",
    x = "Revenue Group",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

# Applied for Financing

-   AIC: 79.96\
-   p \< 0.001\
-   **Insight**: Mid revenue firms (\$25k-\$500k) were less likely to apply. This may reflect barriers to access or low confidence in approval.

------------------------------------------------------------------------

# Model 3: Funding Success (Most/All Received) (Setup)

```{r model3, include=TRUE}
outcome_success <- clean_names %>%
  filter(str_detect(survey_question, "(?i)best outcome across loan")) %>%
  mutate(success = case_when(
    response_option %in% c("Most (51-99%)", "All (100%)") ~ 1,
    TRUE ~ 0
  ),
  yes = round(percent * question_sample_size),
  no = question_sample_size - yes,
  revenue = factor(revenue)) %>%
  group_by(revenue, success) %>%
  summarise(yes = sum(yes), no = sum(no), .groups = "drop")

model_outcome <- glm(cbind(yes, no) ~ revenue, data = outcome_success, family = binomial())
outcome_success$predicted <- predict(model_outcome, type = "response")
```

------------------------------------------------------------------------

# Model 3: Funding Success (Plot)

```{r plot3, echo=FALSE}
ggplot(outcome_success, aes(x = revenue, y = predicted)) +
  geom_col(fill = "#984ea3") +
  coord_flip() +
  labs(
    title = "Predicted Probability of Receiving Most/All Requested Funding",
    x = "Revenue Group",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

---
# Received Most/All Requested Funding 
AIC  695.44 Not significant
Insight Revenue did not predict success in receiving full funding. Other variables (e.g., credit score, lender type) likely matter more.
---

# Model 4: Profit Success (Setup)

```{r model4, include=TRUE}
profit_success <- clean_names %>%
  filter(str_detect(survey_question, "(?i)profitability, end of 2023")) %>%
  mutate(
    success = if_else(response_option == "Operating at a profit", 1, 0),
    yes = round(percent * question_sample_size),
    no = question_sample_size - yes,
    revenue = factor(revenue)
  ) %>%
  group_by(revenue, success) %>%
  summarise(
    yes = sum(yes),
    no = sum(no),
    .groups = "drop"
  )

model_profit <- glm(cbind(yes, no) ~ revenue,
                    data = profit_success,
                    family = binomial())

profit_success$predicted <- predict(model_profit, type = "response")
```

------------------------------------------------------------------------

# Model 4: Profit Success (Plot)

```{r plot4, echo=FALSE}
ggplot(profit_success, aes(x = revenue, y = predicted)) +
  geom_col(fill = "#e41a1c") +
  coord_flip() +
  labs(
    title = "Predicted Probability of Being Profitable (End of 2023)",
    x = "Revenue Group",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

# Profitability (End of 2023)

AIC 23.59

Not significant

Insight Revenue ≠ profit. Some high-revenue businesses were not profitable.

------------------------------------------------------------------------

# Model 5: Condition of Success (Setup)

```{r model5, include=TRUE}
condition_success <- clean_names %>%
  filter(str_detect(survey_question, "(?i)current financial condition")) %>%
  mutate(
    success = case_when(
      response_option %in% c("Good", "Very good", "Excellent") ~ 1,
      TRUE ~ 0
    ),
    yes = round(percent * question_sample_size),
    no = question_sample_size - yes,
    revenue = factor(revenue)
  ) %>%
  group_by(revenue, success) %>%
  summarise(
    yes = sum(yes),
    no = sum(no),
    .groups = "drop"
  )

model_condition <- glm(cbind(yes, no) ~ revenue,
                       data = condition_success,
                       family = binomial())

condition_success$predicted <- predict(model_condition, type = "response")
```

------------------------------------------------------------------------

# Model 5: Condition of Success (Plot)

```{r plot5, echo=FALSE}
ggplot(condition_success, aes(x = revenue, y = predicted)) +
  geom_col(fill = "#ff7f00") +
  coord_flip() +
  labs(
    title = "Predicted Probability of Positive Financial Condition",
    x = "Revenue Group",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

# Positive Financial Condition

AIC: 19.14

Not significant

Insight: Subjective financial wellness wasn’t strongly tied to revenue level.

------------------------------------------------------------------------

# Model 6: Employemt Success (Setup)

```{r model6, include=TRUE}
employment_success <- clean_names %>%
  filter(str_detect(survey_question, "(?i)employment expectations")) %>%
  mutate(
    success = if_else(response_option == "Will increase", 1, 0),
    yes = round(percent * question_sample_size),
    no = question_sample_size - yes,
    revenue = factor(revenue)
  ) %>%
  group_by(revenue, success) %>%
  summarise(
    yes = sum(yes),
    no = sum(no),
    .groups = "drop"
  )

model_employment <- glm(cbind(yes, no) ~ revenue,
                        data = employment_success,
                        family = binomial())

employment_success$predicted <- predict(model_employment, type = "response")
```

------------------------------------------------------------------------

# Model 6: Employment Success (Plot)

```{r plot6, echo=FALSE}
ggplot(employment_success, aes(x = revenue, y = predicted)) +
  geom_col(fill = "#a65628") +
  coord_flip() +
  labs(
    title = "Predicted Probability of Increasing Employment",
    x = "Revenue Group",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

# Employment Growth Expectations

AIC: 4.40

Not significant

Insight: Hiring plans appear strategy-based, not revenue-based.

------------------------------------------------------------------------

# Model 7: Combind Business data (Setup)

```{r model7, include=TRUE}
clean_data <- clean_names %>%
  rename(
    question = survey_question,
    response = response_option,
    sample_size = question_sample_size
  )


business_data <- clean_data %>%
  mutate(
    applications = if_else(response == "Applied for a loan...", 1, 0),
    outcome = if_else(response %in% c("Most (51-99%)", "All (100%)"), 1, 0),
    success = applications + outcome,  # optional: create score
    yes = round(percent * sample_size),
    no = sample_size - yes
  )

model_null <- glm(cbind(yes, no) ~ 1, data = business_data, family = binomial())
model_applications <- glm(cbind(yes, no) ~ applications, data = business_data, family = binomial())
model_combined <- glm(cbind(yes, no) ~ applications + outcome, data = business_data, family = binomial())
```

# Combined Predictors:

Applied for financing

Received most/all funding

AIC: 474,925.9 (lowest of all)

p \< 2.2e-16

Insight: Businesses that both seek and secure funding are more likely to be successful. This model strongly predicts resilience and growth.

------------------------------------------------------------------------

# Summary

Revenue level alone is a weak predictor of success.

Financing behavior particularly the act of seeking and receiving funding is a much stronger indicator.

Policy Implication: Programs aimed at improving capital access for small businesses could support long-term stability and expansion.

# Thank You

Questions?\
LaVelle White\
Data Science \@ Chaminade University
