---
title: "Practice: Non-linear Transformations of Predictors"
author: 
  - "LaVelle White"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/5/25
format: html
editor: visual
theme: spacelab
---

## Non-linear Transformations of Predictors

Using non-linear transformations of predictors, like adding `lstat2` in a regression model, can help capture non-linear relationships between the predictors and the response variable.

Here's why this can be beneficial:

1.  **Improved Fit**: If the relationship between the predictor and the response is not strictly linear, incorporating polynomial terms can improve the model's ability to fit the data.

2.  **Flexibility**: By including terms like `lstat2`, you allow the model to adjust more flexibly to the underlying data patterns.

3.  **Capturing Non-Linear Effects**: In many real-world situations, increases in one variable may have diminishing or increasing effects on another. For instance, the impact of `lstat` on `medv` may not be uniform across the whole range.

4.  **Model Diagnostics**: By examining models with and without the non-linear terms, you can evaluate which model better explains the variability in the response variable, often assessed using criteria like R-squared or AIC.

## Let's try it

Start by loading the `MASS` and `ISLR2` packages, and also `lme4` and `tidyverse`. For this exercise, we'll continue with the `Boston` data.

library(MASS)

library(ISLR2)

library(lme4)

library(tidyverse)

```{r, echo = FALSE, message = FALSE}
# install.packages("ISLR2")
# install.packages("lme4")

library(MASS)
library(ISLR2)
library(lme4)
library(tidyverse)
```

## Boston Housing Data

Along with linear models, the `lm()` function can handle non-linear transformations of predictors too! For instance, if you have a predictor `lstat`, you can create a new predictor `lstat2` using `I(lstat^2)`. We use the `I()` function because the `^` symbol has a special role in formula objects, and wrapping it this way helps us use the standard R method to raise `lstat` to the power of 2. Now, let’s run a regression of `medv` on both `lstat` and `lstat2`!

```{r}
data(Boston) # load the Boston data
```

### Basic linear regression syntax

As review, here is the basic model.

```{r}
# # simple linear regression
# model <- lm(y ~ x, data=df)
# 
# # multiple linear regression
# model <- lm(y ~ x + z, data=df)
# 
# # multiple linear regression with an interaction term
# model <- lm(y ~ x * z, data=df)
# 
# # spelled out/long form of the above. they do the same thing
# model <- lm(y ~ x + z + x:z, data=df)
```

### Visualize first

Make a scatterplot!

```{r}
scatter <- ggplot(Boston, aes(x = lstat, y = medv)) +
  geom_point()
print(scatter)
```

### Fit the model with a quadratic term

```{r}
lm.quad <- lm(medv ~ lstat + I(lstat^2), data = Boston) 
summary(lm.quad)
```

The very small *p*-value for the quadratic term indicates that it significantly enhances the model. To better understand how much better the quadratic fit is compared to the linear fit, we can use `anova()` for a detailed comparison.

```{r}
lm.simple <- lm(medv ~ lstat, data = Boston) # run the simple model
# summary(lm.simple) # we have reviewed versions of this model in previous classes

# compare the two models
anova(lm.simple, lm.quad)
```

`Model 1` represents the linear submodel with just one predictor, `lstat`, while `Model 2` is the larger quadratic model that includes both `lstat` and `lstat2`.

The `anova()` function conducts a hypothesis test to compare these two models. The null hypothesis states that both models fit the data equally well, while the alternative hypothesis suggests that the full model is better.

Here, the *F*-statistic is 135, and the corresponding *p*-value is nearly zero. This strongly indicates that the model with both `lstat` and `lstat2` is much better than the one with only `lstat`. This finding aligns with earlier observations of non-linearity in the relationship between `medv` and `lstat`.

### Visualize again, with the regression lines

```{r}
# linear regression (striaght line)
ggplot(Boston, aes(x = lstat, y = medv)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, color = "blue")
  
```

Build another plot adding the quadratic term).

```{r}
#linear regression (striaght line)
ggplot(Boston, aes(x = lstat, y = medv)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), color = "blue")
```

### Higher order polynomials

To create a cubic fit, we can add a predictor in the form of `I(X^3)`. However, this method can become cumbersome for higher-order polynomials. A more efficient approach is to use the `poly()` function to generate the polynomial directly within `lm()`. For instance, the following command creates a fifth-order polynomial fit:

```{r}
lm.fit5 <- lm(medv ~ poly(lstat, 5), data = Boston)
summary(lm.fit5)
```

### Question

The output here looks a little different from what we have been looking at. See what you can find online about this and try to describe it in your own words:

(Our line begins to curve as the cluster become less dense.)

### Continue exploring the model

Below, compare `lm.fit5` with `lm.quad`. What do you find?

```{r}
anova(lm.fit5, lm.quad)
```

Try a higher order polynomial of your choice and compare with other models you've run. What is the best model for this data? (You can also try a logarithmic transformation using `log()`.)

```{r}
lm.poly3 <- lm(medv ~ poly(crim, 3) + poly(rm, 3) + poly(nox, 3), data=Boston)
summary(lm.poly3)

```

### Your turn!

Continue using the Boston data set for now, but pick a couple other variables. As many as you want.

### Visualize

Make a scatterplot to visualize the relationship between your variables.

```{r}
Boston_long <- tidyr::pivot_longer(Boston, cols = c(crim, rm, nox), 
                                   names_to = "Variable", values_to = "Value")


ggplot(Boston_long, aes(x = Value, y = medv, color = Variable)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = FALSE) +
  labs(title = "Relationship of Crime Rate, Rooms, and Air Pollution with Home Value",
       x = "Predictor Value (Crime Rate, Rooms, or NOx)",
       y = "Median Home Value ($1000s)") +
  theme_minimal()

```

### Transform your predictors

See if you can apply polynomials to find a better fit. Be sure to compare your model with the simple `lm()`.

```{r}
# Simple Linear Model (Baseline)
lm_simple <- lm(medv ~ crim + rm + nox, data = Boston)
summary(lm_simple)


# Polynomial Model (Degree 3)
lm_poly3 <- lm(medv ~ poly(crim, 3) + poly(rm, 3) + poly(nox, 3), data = Boston)
summary(lm_poly3)


# Compare AIC and BIC
AIC(lm_simple, lm_poly3)
BIC(lm_simple, lm_poly3)

```

\

## References

<https://chatgpt.com/share/67a45682-8620-800f-bbf0-a705b500fb2c>
